### 2021-02-18 end_appendix/appendix.tex
Added appendix B

The basic elements of the \ac{GBRT} algorithm are decision trees. These models
create decision rules based on binary splits to predict a target variable $y$
from a set of predictors $\bm{x} = \left( x^{(1)}, x^{(2)}, \ldots, x^{(K)}
\right)$ ($K$ is the number of predictors). These predictors do not need to be
of the same type: \ac{GBRT} allows the simultaneous input of numerical and
categorical variables, which is a great advantage for our use case. There is no
need to encode the categorical predictors in any way. Due to their simple
nature, \acf{ML} models based on decision trees are easy to interpret and
explain but cannot be used to create satisfying predictions for complex
datasets. This issue can be overcome by a technique called \emph{boosting}.
Boosting improves the performance of \emph{weak learners} (in our case decision
trees) by combining a large number of them  \autocite{Freund1996}. The
regression function used to predict $\hat{y} = F(\bm{x})$ can be written as a
linear combination of simple decision trees $h(\bm{x})$
\begin{equation}
  F(\bm{x}) = \sum_{j=0}^{H} \beta_j h(\bm{x}, \bm{\alpha}_j),
  \label{eq:app:b:f_x_sum}
\end{equation}
where $H$ is the total number of decision trees, $\beta_j$ the expansion
coefficient for tree $j$ and $\bm{\alpha}_j$ the hyperparameters for tree $j$.
Using all $N$ training data points $\left\{ \left( \bm{x}_i, y_i \right) \mid m
\in I_N \right\}$ with index set $I_N = \left\{ 1, 2, \ldots, N \right\}$
(\enquote{classic} gradient boosting), the expansion coefficients and
parameters are jointly fitted by minimizing a loss function $L \left( y,
F(\bm{x}) \right)$ in a forward iteration:
\begin{equation}
  \left( \beta_j, \bm{\alpha}_j \right) = \underset{\beta,
    \bm{\alpha}}{\arg\min} \sum_{i=1}^{N} L \left( y_i, F_{j-1}(\bm{x}_i) +
  \beta h(\bm{x}_i, \bm{\alpha}) \right).
  \label{eq:app:b:beta_alpha}
\end{equation}
In practice, this iteration step only uses a randomly selected subsample of the
training data (drawn without replacement), \ie{} the sum does not cover all $N$
training points. Starting with an initial guess $F_0(\bm{x})$, the model is
recursively built by
\begin{equation}
  F_j(\bm{x}) = F_{j-1}(\bm{x}) + \beta_j h(\bm{x}, \bm{\alpha}_j).
  \label{eq:app:b:f_x_recursion}
\end{equation}
The minimization procedure of the loss function (mean square error with
additional sample weights determined by the grid cell areas) is called
\emph{stochastic gradient boosting} and is explained in detail by
\textcite{Friedman2001, Friedman2002}. Fitting the \ac{GBRT} model involves
building the decision trees by splitting the data at points with maximum
information gain. Boosting those simple trees greatly improves the overall
predictive power of the \ac{ML} algorithm: Poorly modeled training points in
the early stages of the algorithm will gradually improve throughout the
training process.

@@

@@

\begin{figure}[p]
  \centering
  \begin{subfigure}[b]{0.39\columnwidth}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s1a.pdf}
    \caption{}
    \label{fig:app:b:co2:a}
  \end{subfigure}
  \begin{subfigure}[b]{0.39\columnwidth}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s1b.pdf}
    \caption{}
    \label{fig:app:b:co2:b}
  \end{subfigure}
  \begin{subfigure}[b]{0.2\columnwidth}
    \raisebox{14.5mm}{\includegraphics[width=\columnwidth]{
      ch06_paper_gpp/figs/s1c.pdf}}
  \end{subfigure}
  \caption{(a) Monthly mean atmospheric \acs{CO2} concentration at \acf{KUM}
    from 1979 to 2019. The thin colored lines show the individual \acs{CMIP}5
    models (emission-driven historical simulations for the years
    \range{1979}{2005} and emission-driven \acs{RCP}8.5 simulations for the
    years \range{2006}{2019}; the latter is not available for HadGEM2-ES). The
    thick black line shows the observations. For the \acs{CMIP}5 models, the
    grid cell closest to \acs{KUM} is considered. The curves show an increase
    of the atmospheric \acs{CO2} concentration superimposed by a pronounced
    seasonal cycle (see \cref{subsec:02:carbon_cycle_perturbations}). (b)
    Annual amplitude of the seasonal cycle of \acs{CO2} (defined as the
    difference between the maximum and the minimum monthly mean atmospheric
    \acs{CO2} concentration for each year) against the annual mean atmospheric
    \acs{CO2} concentration at \acs{KUM}. Colored points show the \acs{CMIP}5
    models (similar time ranges as in (a)) and thick black points show the
    observations. The lines show the corresponding linear regression fits for
    each dataset. The slopes of these linear fits define the sensitivity of
    the seasonal \acs{CO2} cycle amplitude to atmospheric \acs{CO2}
    concentrations, which is used as predictor for the emergent constraint
    step of our approach (step 1). \AdaptedFrom{Schlund2020}.}
  \label{fig:app:b:co2}
\end{figure}

@@

\begin{figure}[p]
  \centering
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5a.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:a}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5b.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:b}
  \end{subfigure}
  \\
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5c.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:c}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5d.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:d}
  \end{subfigure}
  \\
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5e.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:e}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5f.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:f}
  \end{subfigure}
  \caption{Geographical distributions of the \acfp{SPE} of the absolute
    \acf{GPP} at the end of the \nth{21} century in the \acs{RCP}8.5 scenario
    (step 2a) for different statistical models. Details on the calculation of
    the \acs{SPE} are given in \cref{subsec:app:b:xxx}. (a) \acs{CMIP}5
    \acf{MMM}. (b) Rescaled \acs{CMIP}5 \acs{MMM} using \cref{eq:06:y}. (c)
    \Acf{LASSO} model using only the historical \acs{GPP} as single predictor.
    (d) \Acf{GBRT} model using only the historical \acs{GPP} as single
    predictor. (e) \acs{LASSO} model using all predictors. (f) \acs{GBRT}
    model using all predictors. The \acs{SPE} is minimal for the \acs{GBRT}
    model using all predictors. \AdaptedFrom{Schlund2020}.}
  \label{fig:app:b:step2a_results_errors}
\end{figure}

\begin{figure}[p]
  \centering
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s6a.pdf}
    \caption{}
    \label{fig:app:b:step2b_results_errors:a}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s6b.pdf}
    \caption{}
    \label{fig:app:b:step2b_results_errors:b}
  \end{subfigure}
  \\
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s6c.pdf}
    \caption{}
    \label{fig:app:b:step2b_results_errors:c}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s6d.pdf}
    \caption{}
    \label{fig:app:b:step2b_results_errors:d}
  \end{subfigure}
  \caption{Geographical distributions of the \acfp{SPE} of the fractional
    change in \acf{GPP} over the \nth{21} century in the \acs{RCP}8.5 scenario
    (step 2b) for different statistical models. Details on the calculation of
    the \acs{SPE} are given in \cref{subsec:app:b:xxx}. (a) \acs{CMIP}5
    \acf{MMM}. (b) Rescaled \acs{CMIP}5 \acs{MMM} using \cref{eq:06:y}. (c)
    \Acf{LASSO} model. (d) \Acf{GBRT} model. The \acs{SPE} is minimal for the
    \acs{GBRT} model. \AdaptedFrom{Schlund2020}.}
  \label{fig:app:b:step2b_results_errors}
\end{figure}

>>>>

The basic elements of the \ac{GBRT} algorithm are decision trees. These models
create decision rules based on binary splits to predict a target variable $y$
from a set of predictors $\bm{x} = \left( x^{(1)}, x^{(2)}, \ldots, x^{(K)}
\right)$ ($K$ is the number of predictors). These predictors do not need to be
of the same type: \ac{GBRT} allows the simultaneous input of numerical and
categorical variables, which is a great advantage for our use case. There is no
need to encode the categorical predictors in any way. Due to their simple
nature, \acf{ML} models based on decision trees are easy to interpret and
explain but cannot be used to create satisfying predictions for complex
datasets. This issue can be overcome by a technique called \emph{boosting}.
Boosting improves the performance of \emph{weak learners} (in our case decision
trees) by combining a large number of them  \autocite{Freund1996}. The
regression function used to predict $\hat{y} = F(\bm{x})$ can be written as a
linear combination of simple decision trees $h(\bm{x})$
\begin{equation}
  F(\bm{x}) = \sum_{j=0}^{H} \beta_j h(\bm{x}, \bm{\alpha}_j),
  \label{eq:app:b:f_x_sum}
\end{equation}
where $H$ is the total number of decision trees, $\beta_j$ the expansion
coefficient for tree $j$ and $\bm{\alpha}_j$ the hyperparameters for tree $j$.
Using all $N$ training data points $\left\{ \left( \bm{x}_i, y_i \right) \mid i
\in I_N \right\}$ with index set $I_N = \left\{ 1, 2, \ldots, N \right\}$
(\enquote{classic} gradient boosting), the expansion coefficients and tree
parameters are jointly fitted by minimizing a loss function $L \left( y,
F(\bm{x}) \right)$ in a forward iteration:
\begin{equation}
  \left( \beta_j, \bm{\alpha}_j \right) = \underset{\beta,
    \bm{\alpha}}{\arg\min} \sum_{i=1}^{N} L \left( y_i, F_{j-1}(\bm{x}_i) +
  \beta h(\bm{x}_i, \bm{\alpha}) \right).
  \label{eq:app:b:beta_alpha}
\end{equation}
In practice, this iteration step only uses a randomly selected subsample of the
training data (drawn without replacement), \ie{} the sum does not cover all $N$
training points. Starting with an initial guess $F_0(\bm{x})$, the model is
recursively built by
\begin{equation}
  F_j(\bm{x}) = F_{j-1}(\bm{x}) + \beta_j h(\bm{x}, \bm{\alpha}_j).
  \label{eq:app:b:f_x_recursion}
\end{equation}
The minimization procedure of the loss function (mean square error with
additional sample weights determined by the grid cell areas) is called
\emph{stochastic gradient boosting} and is explained in detail by
\textcite{Friedman2001, Friedman2002}. Fitting the \ac{GBRT} model involves
building the decision trees by splitting the data at points with maximum
information gain. Boosting those simple trees greatly improves the overall
predictive power of the \ac{ML} algorithm: Poorly modeled training points in
the early stages of the algorithm will gradually improve throughout the
training process.

@@

\subsection{Evaluation of Prediction Uncertainty}
\label{subsec:app:b:uncertainty}

We estimate the \acf{SPE} of the \ac{GBRT} model itself as the \ac{RMSE} of the
predicted target variable $\hat{y}$ and the corresponding true values $y$ of a
hold-out test dataset, the so-called \emph{\acf{RMSEP}} \autocite{Bishop2006},
which is assumed to be constant for all prediction input points:
\begin{equation}
  \sigma_\mathtxt{GBRT} = \mathtxt{RMSE}(\hat{y}, y).
  \label{eq:app:b:sigma_gbrt}
\end{equation}
For this, we randomly select $25 \unit{\%}$ of the input data prior to
training, so that this part of the data neither enters the training of the
\ac{GBRT} model nor the hyperparameter optimization process. Moreover, the test
dataset allows an assessment of the prediction residuals, which is useful to
detect overfitting (see \cref{fig:app:b:residuals}).

A second source of uncertainty is the error in the rescaling of the target
variable in step 1 of our approach. Analogous to \cref{eq:06:y}, we estimate
this error as
\begin{equation}
  \sigma_{i, \mathtxt{RESC}} = \bar{y}_i \cdot \frac{\sigma_{f'}}{\bar{f}}
  \label{eq:app:b:sigma_rescaling}
\end{equation}
for each prediction input point $i$ ($i$ runs over all grid cells and months).
$\bar{y}_i$ is the \acs{CMIP}5 \acf{MMM} of the target variable at prediction
input point $i$ (step 2a: absolute \ac{GPP} at the end of the \nth{21} century;
step 2b: fractional \ac{GPP} change over the \nth{21} century), $\sigma_{f'}$
the standard error in the global mean fractional \acs{GPP} change given by the
emergent constraint from step 1 and $\bar{f}$ the \acs{CMIP}5 \ac{MMM} global
mean fractional \ac{GPP} change over the \nth{21} century.

The final source of uncertainty is the error of the prediction input data
$\sigma_{ik}$ ($i$ corresponds to the prediction input point and $k$ to the
predictor). These are only available for the FLUXNET-MTE product
\autocite{Jung2011}. To account for this, we use the \acf{LIME} technique
\commentcite{Ribeiro2016}{see \cref{subsec:06:step_2}} to build a local linear
model for every sample point, which yields the linear coefficients $b_{ik}$.
Using error propagation for all predictors and assuming independence of all
individual errors, the \ac{SPE} due to observational uncertainty $\sigma_{i,
  \mathtxt{OBS}}$ then be calculated as
\begin{equation}
  \sigma_{i, \mathtxt{OBS}} = \sum_{k=1}^{K} b_{ik}^2 \sigma_{ik}^2.
  \label{eq:app:b:sigma_obs}
\end{equation}
The total \ac{SPE} at a prediction input point $i$ is the sum of the squared
errors presented above (assuming all of them are independent):
\begin{equation}
  \sigma_i^2 = \sigma_\mathtxt{GBRT}^2 + \sigma_{i, \mathtxt{RESC}}^2 +
  \sigma_{i, \mathtxt{OBS}}^2.
  \label{eq:app:b:sigma_sum}
\end{equation}
The specified error ranges for the \ac{MMM} approaches are calculated in a
similar way: The constant \ac{SPE} per grid cell is estimated by the mean
\ac{RMSEP} given in the leave-one-model-out \ac{CV} experiment (see
\cref{subsec:02:model_weighting}). For the plain \ac{MMM}, this is the only
source of uncertainty. For the rescaled \ac{MMM}, the total error can be
calculated similarly to \cref{eq:app:b:sigma_sum} without the last term
(observational uncertainty).

\subsection{Evaluation of Residuals}
\label{subsec:app:b:residuals}

A convenient way to gain information about statistical models is to analyze the
residuals $\varepsilon_i$, which are defined as the difference between the true
value of the target variable $y_i$ and the predicted value $\hat{y}_i$ at a
sample point $i$ with known ground truth:
\begin{equation}
  \varepsilon_i = y_i - \hat{y}_i.
  \label{eq:app:b:residuals}
\end{equation}
A common way to visualize the residuals is to plot their \acfp{PDF} (see
\cref{fig:app:b:residuals}). The two panels (for steps 2a and 2b) show that the
\ac{ML} model is not overfitting in both cases: The distributions of the
training data and the independent hold-out data are very similar. Moreover, the
\acp{PDF} do not show significant biases, as the residuals are approximately
unbiased (zero mean) for the training and the test dataset. This justifies the
use of the \ac{RMSEP} to estimate the \ac{SPE}, since for unbiased residuals
the \ac{RMSEP} is equal to the standard deviation of the residuals.

@@

\begin{figure}[p]
  \centering
  \begin{subfigure}[b]{0.39\columnwidth}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s1a.pdf}
    \caption{}
    \label{fig:app:b:co2:a}
  \end{subfigure}
  \begin{subfigure}[b]{0.39\columnwidth}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s1b.pdf}
    \caption{}
    \label{fig:app:b:co2:b}
  \end{subfigure}
  \begin{subfigure}[b]{0.2\columnwidth}
    \raisebox{14.5mm}{\includegraphics[width=\columnwidth]{
      ch06_paper_gpp/figs/s1c.pdf}}
  \end{subfigure}
  \caption{(a) Monthly mean atmospheric \acs{CO2} concentration at \acf{KUM}
    from 1979 to 2019. The thin colored lines show the individual \acs{CMIP}5
    models (emission-driven historical simulations for the years
    \range{1979}{2005} and emission-driven \acs{RCP}8.5 simulations for the
    years \range{2006}{2019}; the latter is not available for HadGEM2-ES). The
    thick black line shows the observations. For the \acs{CMIP}5 models, the
    grid cell closest to \acs{KUM} is considered. The curves show an increase
    of the atmospheric \acs{CO2} concentration superimposed by a pronounced
    seasonal cycle (see \cref{subsec:02:carbon_cycle_perturbations}). (b)
    Annual amplitude of the seasonal cycle of \acs{CO2} (defined as the
    difference between the maximum and the minimum monthly mean atmospheric
    \acs{CO2} concentration for each year) against the annual mean atmospheric
    \acs{CO2} concentration at \acs{KUM}. Colored points show the \acs{CMIP}5
    models (similar time ranges as in (a)), and thick black points show the
    observations. The lines show the corresponding linear regression fits for
    each dataset. The slopes of these linear fits define the sensitivity of
    the seasonal \acs{CO2} cycle amplitude to atmospheric \acs{CO2}
    concentrations, which is used as predictor for the emergent constraint
    step of our approach (step 1). \AdaptedFrom{Schlund2020}.}
  \label{fig:app:b:co2}
\end{figure}

@@

\begin{figure}[p]
  \centering
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5a.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:a}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5b.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:b}
  \end{subfigure}
  \\
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5c.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:c}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5d.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:d}
  \end{subfigure}
  \\
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5e.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:e}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s5f.pdf}
    \caption{}
    \label{fig:app:b:step2a_results_errors:f}
  \end{subfigure}
  \caption{Geographical distributions of the \acfp{SPE} of the absolute
    \acf{GPP} at the end of the \nth{21} century in the \acs{RCP}8.5 scenario
    (step 2a) for different statistical models. Details on the calculation of
    the \acs{SPE} are given in \cref{subsec:app:b:uncertainty}. (a) \acs{CMIP}5
    \acf{MMM}. (b) Rescaled \acs{CMIP}5 \acs{MMM} using \cref{eq:06:y}. (c)
    \Acf{LASSO} model using only the historical \acs{GPP} as single predictor.
    (d) \Acf{GBRT} model using only the historical \acs{GPP} as single
    predictor. (e) \acs{LASSO} model using all predictors. (f) \acs{GBRT}
    model using all predictors. The \acs{SPE} is minimal for the \acs{GBRT}
    model using all predictors. \AdaptedFrom{Schlund2020}.}
  \label{fig:app:b:step2a_results_errors}
\end{figure}

\begin{figure}[p]
  \centering
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s6a.pdf}
    \caption{}
    \label{fig:app:b:step2b_results_errors:a}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s6b.pdf}
    \caption{}
    \label{fig:app:b:step2b_results_errors:b}
  \end{subfigure}
  \\
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s6c.pdf}
    \caption{}
    \label{fig:app:b:step2b_results_errors:c}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch06_paper_gpp/figs/s6d.pdf}
    \caption{}
    \label{fig:app:b:step2b_results_errors:d}
  \end{subfigure}
  \caption{Geographical distributions of the \acfp{SPE} of the fractional
    change in \acf{GPP} over the \nth{21} century in the \acs{RCP}8.5 scenario
    (step 2b) for different statistical models. Details on the calculation of
    the \acs{SPE} are given in \cref{subsec:app:b:uncertainty}. (a) \acs{CMIP}5
    \acf{MMM}. (b) Rescaled \acs{CMIP}5 \acs{MMM} using \cref{eq:06:y}. (c)
    \Acf{LASSO} model. (d) \Acf{GBRT} model. The \acs{SPE} is minimal for the
    \acs{GBRT} model. \AdaptedFrom{Schlund2020}.}
  \label{fig:app:b:step2b_results_errors}
\end{figure}
